% !TEX root = ../bigmpi.tex

\section{Conclusions and Future Work}

The MPI Forum issued a challenge to consumers of MPI: ``prove to us derived
datatypes are insufficient''.  In a time where 64 bit systems are widespread,
but C integer types remain 32 bits, describing large memory or file requests
will more frequently require the special handling BigMPI provides.  The
exercise has also revealed several difficulties in the standard, particularly
in the area of user-defined reductions.

\subsection{Suggestions for MPI-4}

% (1) 34+338 (+339?) required for sane reductions.
Tickets 34~\cite{ticket34} and 338~\cite{ticket338}
% https://svn.mpi-forum.org/trac/mpi-forum-web/ticket/338
would allow BigMPI (and other users such as PETSc) to leverage ``accumulate-style
behavior'':  the built-in operations can work on user-defined
datatypes, but on a element-wise basis.

Ticket 339~\cite{ticket339} % https://svn.mpi-forum.org/trac/mpi-forum-web/ticket/339
(``User-defined op with derived datatypes yields space-inefficient reduce'')

% (2) 423 is sugar but it greatly reduces user pain and is trivial to implement.
Ticket 423~\cite{ticket423} % https://svn.mpi-forum.org/trac/mpi-forum-web/ticket/423
(``add MPI\_Type\_contiguous\_x'') is syntactic sugar, but it will go a long way towards reducing users' friction when dealing with large counts.  As is evidenced by BigMPI, the change is straightforward to implement.

%430 is required to support nonblocking v-collectives in a straightforward way.
When applying BigMPI's large-count strategy to the v-collectives, the (counts[], type) description has to be mapped to (newcount[], newtypes[]),
and that in turn requires the w-variants.
Ticket 430~\cite{ticket430}
% https://svn.mpi-forum.org/trac/mpi-forum-web/ticket/430
(``large-count v-collectives'') would provide a large-count v-collective and would avoid the need for big temporary memory allocations.

The implementation of non-blocking collectives in BigMPI requires improved generalized requests.
%MPICH-style generalized requests required (https://svn.mpi-forum.org/trac/mpi-forum-web/wiki/Proposal should be ticket-ized)
Ticket 457~\cite{ticket457} % https://svn.mpi-forum.org/trac/mpi-forum-web/ticket/457
(``expose progress in generalized requests'') would provide sufficient improvement.

\subsection{Suggestions for users}

% jeff: i don't know what you had in mind here...
Our foremost suggestion to users would be to participate in the MPI Forum.  The
MPI Forum is faced with the difficult task of defining a standard that can be
both used and implemented.  Ideally, one would like to be able to say ``easily
used'' and ``easily implemented'', though tradeoffs are sometimes required.
Participation in the MPI Forum, currently dominated by implementors, can
provide a voice to those whose perspectives are less well heard.

Our second suggestion to users is actually a request: please help provide more
test cases.  Coverage of large count workloads remains sparse.  The sizes we
think of as ``unimaginably large'' today will simply be ``large'' tomorrow.
