% !TEX root = ../bigmpi.tex

\subsection{Limitations}

BigMPI does not support the full
range of \texttt{MPI\_Count}, but rather only the range of the
address space (i.e., \texttt{size\_t} and \texttt{MPI\_Aint}), since buffers
larger than the address space are difficult to allocate.

BigMPI supports  only  built-in datatypes.  
Code already using derived-datatypes should already be able 
to handle large counts without BigMPI\@.
However, see Section~\ref{sec:romio_typeproc} 
for an example of HINDEXED not being sufficient.

Support for \texttt{MPI\_IN\_PLACE} is not 
%GWP - Well, yes, if it is impossible, I guess it won't be done -- could you give a better example?
implemented in some cases (e.g., where it is impossible) and 
implemented inefficiently (i.e., via a buffer copy) in others.
Using \texttt{MPI\_IN\_PLACE} is discouraged at the present time
although we expect that it will be supported efficiently whenever
possible in the release version of BigMPI.

BigMPI requires C99.  Fifteen years is more than enough time for compiler 
implementors interested in supporting ISO languages to provide a C99
compiler.

The MPI-3 standard supports language bindings for C and
Fortran---the latter via three different mechanisms
(\texttt{mpif.h}, \texttt{use mpi} and \texttt{use mpi\_f08}).
Currently, BigMPI provides only a C interface, but a Fortran 2003
interface to the C API via \texttt{ISO\_C\_BINDING} is planned.
We expect that C++ programmers will be able to use the
C interface and can implement wrappers consistent with their own
style of C++ programming.
